Online {\IterComp}

### 2017-09-14 infra.tex
itercomp overview

\section{Online {\IterComp} Infrastructure} \label{sec:oic-infra}

Although {\itercomp} had been originally proposed as an \textit{offline} optimization strategy, it can also be adapted to work in online scenarios.
Instead of selecting the best optimization sequence as part of the development time (pre-shipping) of a program, a first version of the program is shipped together with an {\itercomp} mechanism.
In the online scenario, the program is shipped with an initial optimization sequence and different optimization sequences are evaluated as the end-user executes the program.
This optimization strategy is also known as idle-time optimization, as the re-compilation happens between runs of the program.

LLVM is particularly suitable for iterative compilation as it makes possible to cache a pre-compiled, but still unoptimized, version of the input program in the bitcode format of the LLVM IR.
This caching allows to speedup the time required for re-compilation as it is able to bypass the frontend phase.
If re-compilation time is critical, it would also be possible to keep only the hot portion of the code in the LLVM bitcode format, while the remaining portion of the code is already compiled to the final object code.
However, this is out of the scope of this paper and we always re-compile the whole program.

\begin{figure}[htb]
    \centering
    \includegraphics[width=\linewidth]{figs/infra-diagram}
    \caption{Overview of the execution engine for applying {\itercomp}.}
    \label{fig:infra-diagram}
\end{figure}
%\textbf{Describe the process step by step.}

Figure~\ref{fig:infra-diagram} shows an overview of the infrastructure required for applying online {\itercomp}
using WP as the metric of choice for evaluating optimization sequences.
The online {\itercomp} follows as described bellow:
\begin{enumerate}
\item The program is pre-compiled to the LLVM bitcode format without optimization.
\item The unoptimized program is instrumented for work profiling.
\item Execution-based optimization search:
 \begin{enumerate}
   \item The current optimization sequence is used for the re-compilation of the program.
   \item The program is executed with any input provided by the end-user.
         During the execution of the program, wall-clock time and the work metric are recorded by profiling instrumentation.
   \item If the recorded profiling for the current optimization can be used to compute an average performance measurement within a small confidence interval,
         then a new optimization sequence is generated.
         Otherwise, the same optimization sequence is used for the next execution.
 \end{enumerate}
\end{enumerate}

In this work, we focus mainly on the highlighted components.
The \textit{Work Instrumentation} phase focuses on providing a low-overhead instrumentation for profiling the work metric.
The instrumentation consists of adding a global counter to the program which is used to accumulate the amount of work computed during the program's execution, using the cost model of the instruction set.
A detailed description of the work instrumentation is presented in Section~\ref{chap:instr}.

Notice how the \textit{Work Instrumentation} phase is executed before performing any optimization to the program.
It is critical so that the work profiling always measure the same amount of work for a given input, regardless of the optimization sequence applied to the program.
Because the instrumentation is performed before optimising the program, it means that the work profiling derives the linear expression defined in Equation~\ref{eq:linear-work-expression}
based on the unoptimized program.
In other words, the basic blocks and the number of occurrences of the instruction in the basic blocks reflect the unoptimized program.
This particular sequence of compilation guarantees that the amount of work must is only input dependent, but consistent between different optimization sequences.

As the name suggests, the component called \textit{Optimization Selector} is responsible for selecting which optimization sequence to use for the next execution of the program.
It can either keep the same optimization sequence used in the previous execution or start monitoring the performance of a new optimization sequence.
An optimization sequence can be kept for multiple executions of the program in order to gather enough measurement to compute an average performance with small statistical deviations, i.e., for which the confidence interval has a small range.
We call by \textit{Input-Window Size} the number of executions performed using the same optimization sequence.

\subsection{Real Online Scenarios}

In most online scenarios, it is common for periods of peak usage and idle periods.
For example, mobile devices are usually intensely used during the day, with some idle periods usually when the battery is being re-charged~\citep{mpeis16}.
Similarly, many authors have also considered peak and idle (or underutilized) periods in the context of data centres~\citep{armbrust10,chen12b}.

The proposed infrastructure is very well suited for these real online scenarios\footnote{
However, if idle time is almost non-existent, the proposed infrastructure can still be used by re-compiling the program with a different optimization while multiple runs of the program are being executed.}.
In particular, periods of peak usage could be used to monitor multiple runs of the program using the same optimization sequence, collecting the work profiling and measuring its execution time,
while periods of idleness or underutilisation could be leveraged to use the profiling statistics for selecting better optimizations and re-compiling the program.

>>>>

\section{Online {\IterComp} Overview} \label{sec:oic-infra}

Although {\itercomp} had been originally proposed as an \textit{offline} optimization strategy, it can be adapted to work in
\textit{online} scenarios. Instead of selecting the best optimization sequence at development time, a first version of the program is
shipped bundled with the {\itercomp} mechanism. Our mechanism produces new application binaries using different optimization sequences.
When the user starts the application, we load and execute one of these binaries to evaluate the corresponding sequence online. Our
mechanism operates on the application's Intermediate Representation (IR) produced by LLVM. This saves us the effort of repeatedly parsing
the code through the frontend and removes the need for shipping the full application code.

\begin{figure}[htb]
    \centering
    \includegraphics[width=\linewidth]{figs/infra-diagram}
    \caption{Overview of the online \itercomp mechanism.}
    \label{fig:infra-diagram}
\end{figure}

Figure~\ref{fig:infra-diagram} shows an overview of how our online {\itercomp} mechanism works. On the developer's side, the program is
pre-compiled to IR without optimizations and is instrumented to measure the amount of work performed and the wall-clock time. On the user's
side, we select an optimization sequence and then compile the instrumented IR to binary using this sequence. Next time the user runs the
application, we measure the work and the execution time. We accumulate multiple such measurements until we can estimate the efficiency
metric P of this optimization sequence with high confidence. Finally, we use this information to help us select the next optimization
sequence.

This work focuses on the highlighted components of Figure~\ref{fig:infra-diagram}. The first one, \textit{Work Instrumentation} allows
measuring the amount of work performed. Conceptually, we use the cost model of Section~\ref{sec:metric} to estimate a cost for each basic
block of the IR and then add instrumentation to increase a global work counter by that cost every time a basic block is executed. The
measured work should depend only on the program and the input, not the applied optimizations, so we instrument the unoptimized IR before
any optimizations modify it. We present a detailed description of work instrumentation and how to keep it lightweight in
Section~\ref{chap:instr}.

\textit{Optimization Selector} is responsible for selecting which optimization sequence to use for the next execution of the program. Each
sequence is used for multiple executions of the program in order to gather enough work efficiency measurements. When we are able to
estimate this work efficiency with a confidence interval narrower than a certain threshold, we can start evaluating a different
optimization sequence. We call the number of executions performed using the same optimization sequence the \textit{Input-Window Size}.

\FIXME{WHAT IS THE OPTIMIZATION EVALUATOR? WHY IS IT HIGHLIGHTED? DO WE HAVE ANYTHING TO SAY ABOUT IT?}

\subsection{Efficient Optimization Selection and Compilation}

Most devices have distinct idle and peak usage periods. For example, mobile devices are used intensely during the day and not at all while
the user is sleeping and the battery is charging~\citep{mpeis16}. Many authors have identified similar idle and peak usage periods in the
context of data centers~\citep{armbrust10,chen12b}. The proposed \itercomp infrastructure is very well suited to such scenarios. In
particular, we can use periods of idleness or underutilization to select the next few optimization sequences to try and compile the
corresponding binaries. This approach effectively hides the compilation overhead from the user. During peak periods, we only have to select
which of the existing binaries to evaluate which introduces almost no overhead.
